---
title: "Assignment 5: Data Visualization"
author: "Julia Kagiliery"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(here)
library(cowplot)

print(getwd())

NTL_LTER_Lake_Chemistry <- read.csv(here("Data/Processed/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

NEON_NIWO_Litter <- read.csv(here("Data/Processed/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)

#2 

NEON_NIWO_Litter <- NEON_NIWO_Litter |> 
  mutate(collectDate = ymd(collectDate))

NTL_LTER_Lake_Chemistry <- NTL_LTER_Lake_Chemistry |> 
  mutate(sampledate = ymd(sampledate))

typeof(NTL_LTER_Lake_Chemistry$sampledate)
typeof(NEON_NIWO_Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

library(ggridges)
library(viridis)
library(RColorBrewer)
library(colormap)
library(ggthemes)


pink_palette <- c("#FFC0CB", "#FF69B4", "#FF1493", "#DB7093", "#C71585")  # Light to dark pink shades

pink_theme <- theme_base() +
  theme(
    line = element_line(color = 'hotpink', linewidth = 1.5),
    legend.title = element_text(color = 'deeppink', face = "italic"), 
    legend.text = element_text(color = 'darkgrey'),  # Grey legend text
    plot.background = element_rect(color = 'pink'),
    panel.grid.major = element_line(color = 'lightpink', linewidth = .5),
    panel.grid.minor = element_line(color = 'lightpink', linewidth = 0.25),
    axis.title.x = element_text(color = 'deeppink', size = 14),
    axis.title.y = element_text(color = 'deeppink', size = 14),
    plot.title = element_text(color = 'deeppink', size = 16, face = "bold"),
    axis.text.x = element_text(color = 'darkgrey'),  # Grey axis tick labels (x-axis)
    axis.text.y = element_text(color = 'darkgrey'),  # Grey axis tick labels (y-axis)
    strip.text = element_text(color = 'darkgrey'),  # Grey facet labels (if any)
    plot.subtitle = element_text(color = 'darkgrey'),  # Grey subtitle (if any)
    plot.caption = element_text(color = 'darkgrey')  # Grey caption (if any)
  )

# Set default theme
theme_set(pink_theme)


theme_set(pink_theme)

#could not fugure out how to add this above but this function just means that when I use colors to indicate groupings, they will be pink
scale_color_pink <- function() {
  scale_color_manual(values = pink_palette)
}


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
NTL_LTER_Lake_Chemistry |>
  ggplot(aes(x = tp_ug, y = po4, color = as.factor(lakename))) + 
  geom_point() +
    scale_color_pink() +
  ylim(0,50) +
  labs(
    x = "Total Phosphorus (ug)",
    y = "Phosphate (ug)",
    color = "Lake Name",
    title = "Phosphorus v. Phospahte Levels at Various Lakes"
  ) +
  geom_smooth(
    method = lm, 
    se = FALSE
  )


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
TempPlot <- NTL_LTER_Lake_Chemistry |>
  ggplot(aes(y = temperature_C, color = as.factor(lakename))) +
  geom_boxplot(aes(x = factor(month, levels = 1:12, labels = month.abb)), fill = "lightgray", size = 1.5) +  # Fix the factor conversion for month
  scale_color_pink() +
  labs(
    x = element_blank(),
    y = "Temperature (°C)",
    color = "Lake Name",
    title = "Temperature Through the Year"
  )+
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

PhosPlot <- NTL_LTER_Lake_Chemistry |>
  ggplot(aes(y = tp_ug, color = as.factor(lakename))) +
  geom_boxplot(aes(x = factor(month, levels = 1:12, labels = month.abb)), fill = "lightgray", size = 1.5) +  # Fix the factor conversion for month
  scale_color_pink() +
  labs(
    x = element_blank(),
    y = "Total Phosphorus (ug)",
    color = "Lake Name",
    title = "Total Phosphorus Through the Year"
  ) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

NitrogenPlot <- NTL_LTER_Lake_Chemistry |>
  ggplot(aes(y = tn_ug, color = as.factor(lakename))) +
  geom_boxplot(aes(x = factor(month, levels = 1:12, labels = month.abb)), fill = "lightgray", size = 1.5) +  # Fix the factor conversion for month
  scale_color_pink() +
  labs(
    x = "Month",
    y = "Total Nitroge (ug)",
    color = "Lake Name",
    title = "Total Nitrogen Through the Year"
  ) +
    theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))


plot_grid(TempPlot, PhosPlot, NitrogenPlot, nrow = 3,  align = "h", axis = "b")

```



Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: