---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Julia Kagiliery"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
toc: TRUE
---

\listoffigures

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse) 
library(agricolae)
library(dplyr)
library(here)
library(ggplot2)
library(corrplot)
library(ggthemes)
library(crayon)
library(lubridate)

getwd()

LakeChemistry <- read.csv(
  here("Data", "Raw","NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
  stringsAsFactors = TRUE) 

LakeChemistry <- LakeChemistry |>
  mutate(sampledate = mdy(sampledate))

#2

bluetheme <- theme_base() +
  theme(
    line = element_line(color = 'navy', linewidth = 1.5),
    legend.title = element_text(color = 'navy', face = "italic"),
    legend.text = element_text(color = 'steelblue'),
    plot.background = element_rect(color = 'grey'),
    panel.grid.major = element_line(color = 'skyblue', linewidth = .5),
    panel.grid.minor = element_line(color = 'skyblue', linewidth = 0.25),
    axis.title.x = element_text(color = 'steelblue', size = 14),
    axis.title.y = element_text(color = 'steelblue', size = 14),
    plot.title = element_text(
      color = 'steelblue',
      face = "bold"
    ),
    axis.text.x = element_text(color = 'darkgrey'),
    axis.text.y = element_text(color = 'darkgrey'),
    strip.text = element_text(color = 'darkgrey'),
  )

# default 
theme_set(bluetheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temeprature in July does not change with depth across all lakes. 
Ha: The mean lake temeprature in July does change with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, fig.cap='Scatter plot of Temperature of Lakes At Various Depths'}
#4
LakeChemistryClean <- LakeChemistry |>
  filter(month(sampledate) == 7) |>
  select(lakename, year4, daynum, depth, temperature_C) |>
  drop_na()

#5
LakeChemistryClean |>
  ggplot(aes(x = depth, y = temperature_C)) +
  geom_point(color = "navy") +
  geom_smooth(method = "lm", color = "lightpink") +
  labs(
    title = "Temperature of Lakes At Various Depths",
    x = "Depth of Lake (m)",
    y = expression(Temperature~(degree*C))
  ) +
  ylim(0,35) +
  geom_text(x=10 ,y=32, 
label= "Temperature = -1.94621(Depth) + 21.95597 ", size=5, color ="lightpink") # comes from model built in question 7
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: There appears to be a inverse relationship between lake depth and temperature such that increasing depth of the lake has a lower temperature. Given that residuals are all possitive past 10 meters (but our residual values seem to be more negative around 5 meters), I question the validity of a linear model. The possible excessive positive residuals would skew the distribution to the right. Residuals should appear random in a well fit model, but our errors seem to follow a patern (i.e. they may not be normally disttibuted)  and hence I imagine that maybe a logirithmic model might fit better. However, it can be hard to tell this via visual inspection and we should turn to quantitative measuremets and a better analysis of residuals. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
TempDepthModel <- lm(temperature_C~ depth, data = LakeChemistryClean)
print(summary(TempDepthModel))
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The R squared value is 0.7387 for this model so  around 74% of change in temeprature can be explainned by a chnage in depth. Given a P value less than, 0.05, we can say that this is a statistically significat model on 9,726 degrees of freedom. From this, for each meter of depth, we expect that that the temeprature will decrease by a factor of around 1.95 degrees. We expect the surface temperture (depth = 0) to be around 22 degrees C.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
TempDepthModelAIC <- lm(temperature_C~ year4 + daynum + depth, data = LakeChemistryClean)
step(TempDepthModelAIC)

#10
TempDepthModelMultiReg <- lm(temperature_C~ year4 + daynum + depth, data = LakeChemistryClean)
print(summary(TempDepthModelMultiReg))

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: We use all the explanatory variables tested in the model and manage to explain just over 74% of variability in temeprature which is only marginally better than just using depth. Though this is a better model, it is not much better so we should ask if collecting and storing additional data is even worth is for such minor imporvement (should we pay someone to record the date at these lakes for a difference of around 0.4% more explanaituon?)


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

LakeDiffernceLinear<-lm(temperature_C~ lakename, data=LakeChemistryClean)
print(summary(LakeDiffernceLinear))

LakeDiffernceAnova<-aov(temperature_C~ lakename, data=LakeChemistryClean)
print(summary(LakeDiffernceAnova))

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: For the linear models, we see that eacvh lake name has a statistcially significant coeffficent (P value less than 0.05) which suggests that the lakes all have different equations.  For the ANOVA test, the F-value (which discusses inter v intra lake varaitions) of 50 with 8 degrees of freedom and a p-value under 0.05 provides  evidence to reject the null hypothesis and say that the mean temperatures of the lakes are different.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, , fig.cap='Temperature of Lakes At Various Depths by Lake'}
#14.
LakeChemistryClean |>
  ggplot(aes(x = depth, y=temperature_C, color = lakename))+
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Lake Depth and Temperature",
    x = "Depth of Lake (m)",
    y = expression(Temperature~(degree*C)), 
  ) + 
  ylim(0,35) +
scale_color_viridis_d(option = "D", guide = "none") + #deafult colors were ugly
  facet_wrap(~lakename)
```

```{r scatterplot.3, , fig.cap='Temperature of Lakes At Various Depths Drawn by Lake'}
LakeChemistryClean |>
  ggplot(aes(x = depth, y=temperature_C, color = lakename))+
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Lake Depth and Temperature",
    x = "Depth of Lake (m)",
    y = expression(Temperature~(degree*C)), 
    color = "Lake Name"
  ) + 
  ylim(0,35) +
scale_color_viridis_d(option = "D") #deafult colors were ugly
```


15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(LakeDiffernceAnova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:Statistcially, Peter Lake, Paul Lake (p = 0.224), and Ward lake (P=0.783) all have the same mean. No lake has a mean temp statitsically different from all other lakes, 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Two-sample T test would work.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
LakeChemistryCleanCLWL <- LakeChemistryClean |>
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

TTestCLWL<-t.test(temperature_C~ lakename, data=LakeChemistryCleanCLWL)
print(TTestCLWL)

```

>Answer: We can't reject the null the null hypthesis because our p value is not less than 0.05. Furthermore, our 95% confidence interval includes 0 so we cannot be sure that the difference between the means is not = 0. This is somewhat aligned with my answer to 16 which reports the difference between the two lakes as -0.89 but this difference is not signficant!
